Using an iterated block particle filter via \pkg{spatPomp}

### 2023-06-05 ms.Rnw
diagnostics section added

\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{url} 
\newcommand\eic[1]{{\textcolor{orange}{#1}}}
\newcommand\pnc[1]{{\textcolor{blue}{#1}}}
\usepackage{fullpage}
\bibliographystyle{apalike}

@@

The IBPF algorithm studied by \citet{ning23-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw23arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
We also test \code{ibpf} on simulated data for a measles transmission model.
In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file.

\section{Correlated Gaussian random walks} \label{sec:bm}

@@

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{url} 
\newcommand\eic[1]{{\textcolor{orange}{#1}}}
\newcommand\pnc[1]{{\textcolor{blue}{#1}}}
\usepackage{fullpage}
\bibliographystyle{apalike}

@@

\begin{abstract}
The IBPF algorithm studied by \citet{ning23-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw23arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
We also test \code{ibpf} on simulated data for a measles transmission model.
In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file. 
\end{abstract}

\tableofcontents

\newpage

\section{A toy example: Correlated Gaussian random walks} \label{sec:bm}

@@

\section{Diagnostics}

<<diagnostics-setup,eval=T,echo=F>>=
diagnostics_dir <- paste0("diagnostics_",i,"/")
if(!dir.exists(diagnostics_dir)) dir.create(diagnostics_dir)
@

If likelihood maximization appears to be failing, or has unacceptably high Monte Carlo variability, one wishes to diagnose the cause, or causes.
Even if inference appears to be operating satisfactorily, it is good practice to see whether there are some data points for which the model is inappropriate, or some parameters for which optimization is less stable.
It may be interesting to know if less Monte Carlo effort would be sufficient, or if more would be useful.
For all these questions, a starting point is exploratory analysis of diagnostic data produced by \code{ibpf} and \code{bpfilter}.

The success of an iterated filter for likelihood maximization depends on the success of a single pass of that filter for likelihood evaluation.
A useful diagnostic quantity for the particle filter is the effective sample size (ESS) \citep{douce11,liu01}.
In the notation of \citep{king16}, the filter observation weight for particle $j$ at time $n$ is $w(n,j)=f_{Y_n|X_n}(y_n^*|X^P_{n,j};\theta)$, the normal weight is $\tilde{w}(n,j)=w(n,j)\big/ \sum_{k=1}^Jw(n,k)$, and the effective sample size is
\begin{equation}
\mathrm{ESS}_n = \left(\sum_{j=1}^J \tilde{w}(n,j)^2\right)^{-1}.
\end{equation}
ESS is generally motivated as an approximation to the equivalent number of independent samples from the filtering distribution. 
In the context of likelihood-based inference, it is convenient to think of $\mathrm{ESS}^{-1}_n$ as an approximation to the variance of the conditional log-likelihood estimate,
\begin{equation}
\hat \ell_n = \log \left( \frac{1}{J} \sum_{j=1}^J w(n,j) \right).
\end{equation}
However, ubiquitous multi-core computation permits us to calculate this variance directly, by Monte Carlo replication.
This bypasses consideration of when ESS is, and is not, a good estimator. 
Also, the $R$ Monte Carlo replicates, $\hat\ell_n^{(r)}$, $r=1,\dots,R$, give rise to an improved estimate of the conditional log-likelihood,
\begin{equation}
\bar \ell_n = \frac{1}{R} \sum_{r=1}^R \hat\ell_n^{(r)},
\end{equation}
together with an uncertainty estimate via the central limit theorem.

<<pfilter-test-london,eval=T,echo=F,fig.height=6, fig.width=6, out.width="4.5in", fig.cap = "Comparison of effective sample size and conditional log-likelihood variance for measles in London">>=
london <- he10(U=1)
london_diag <- bake(file=paste0(diagnostics_dir,"london.rds"),seed=230604, {
  pf_list <- foreach(rep=1:switch(i,6,100)) %dopar% pfilter(london,Np=switch(i,10,5000))
  ess=sapply(pf_list,function(x)x@eff.sample.size)
  ll=sapply(pf_list,function(x)x@cond.logLik)
  list(mess=apply(ess,1,mean),cllv=apply(ll,1,var),mll=apply(ll,1,mean))
})
london_ess <- london_diag$mess; london_cllv <- london_diag$cllv; londom_mll <- london_diag$mll
xy_range <- range(c(1/london_ess,london_cllv))
plot(x=london_cllv,y=1/london_ess,log="xy",ylab="1/ESS",xlab="Conditional log-likelihood variance",
  xlim=xy_range, ylim = xy_range)
abline(a=0,b=1)
@

Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
We see that both measures agree on the most problematic observation.
However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution, and so we prefer to work with CLLV.
CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
Using the notation of \citep{asfaw23}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
\begin{equation}
w^j_{k,n} = \prod_{u \in {\mathcal B}_k}
            f_{Y_{n,n}|X_{u,n}}
            \big(
              y^*_{u,n}\given X^{P,j}_{u,n} \giventh \theta
            \big).
\end{equation}
The BCLL is
\begin{equation}
\hat \ell_{k,n} = \frac{1}{J}\sum_{j=1}^J w^j_{k,n},
\end{equation}
and the block approximation to the conditional likelihood of $y^*_1,\dots,y^*_{U}$ is
\begin{equation}
\hat\ell_n = \sum_{k=1}^K \hat \ell_{k,n}.
\end{equation}
For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
<<bpfilter-diag-code,eval=F,echo=T>>=
bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {
  bpfilter(he10_model,Np=switch(i,10,2000),block_size=1)
}
bpf_ll  <- sapply(bpf_list,function(x)x@block.cond.loglik)
dim(bpf_ll) <- c(dim(bpf_list[[1]]@block.cond.loglik),length(bpf_list))
bcll <- apply(bpf_ll,c(1,2),mean)
bcllv <- apply(bpf_ll,c(1,2),var)
@

Once we have identified that an observation is hard to filter, there are various classes of explanation: (i) the data point is an outlier, which any reasonable model would struggle to explain; (ii) the model is poor at that point; (iii) there is no problem with the model or data, but the system happens to require a high Monte Carlo effort at that point.
Neither ESS nor CLLV can distinguish these alternatives.
It can be helpful to compare the model fit against a simple statistical model, to provide an objective benchmark.
Measles case counts are an example of a population system exhibiting exponential growth and decay, and in such cases a log-scale autoregressive moving average (log-ARMA) model can provide an appropriate benchmark.
The function \code{spatPomp::arma\_benchmark} fits this model for each unit of a \code{spatPomp} object and provides the unit conditional log-likelihood values for each time point.
Here, blocks are taken to be units, so we define the log-likelihood anomaly as the difference between the BCCL and the unit benchmark,
<<anomaly,echo=T,eval=F>>=
benchmark <- arma_benchmark(he10_model)
anomaly <- bcll - benchmark$cond
@
Unlike BCLL, both log-likelihood anomaly and BCLLV have the convenient property that they do not depend on the scale of the data, or the units in which it is measured.

Figure~\ref{fig:bpfilter-diagnostics} investigates two units---London and Birmingham---via BCLLV and log-likelihood anomaly.
We see that the model used here fits London well and Birmingham poorly.
Some time points have very high Monte Carlo likelihood estimate variance, and those points also have poor likelihood compared to the benchmark.
One observation for London has a poor log-likelihood anomaly but a decent BCLLV.
For Birmingham, both anomaly and BCLLV lead the same conclusion that the model is unsatisfactory, at these default initial parameter values.
Here, our goal is only to check methods on simulated data, but if we wanted to fit the model to data, we should confirm whether the diagnostics from the final fitted model have improved.

<<bpfilter-diagnostics,eval=T,echo=F,fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Block particle filter diagnostics for London (A,C) and Birmingham (B,D).">>=
bpf_diag <- bake(file=paste0(diagnostics_dir,"bpfilter.rds"),seed=230604, {
<<bpfilter-diag-code>>
  list(bcll=bcll,bcllv=bcllv)
})
bcll <- bpf_diag$bcll; bcllv <- bpf_diag$bcllv
<<anomaly>>
par(mfrow=c(2,2))
par(mai=c(1,0.8,0.1,0.1))
plot(x=bcllv[1,],y=anomaly[1,],log="x",ylab="anomaly",xlab="Conditional log-likelihood variance")
mtext("A",side=3,line=-1,adj=-0.24,cex=1.5)
plot(x=bcllv[2,],y=anomaly[2,],log="x",ylab="anomaly",xlab="Conditional log-likelihood variance")
mtext("B",side=3,line=-1,adj=-0.24,cex=1.5)
plot(y=anomaly[1,],x=time(he10_model),ylab="anomaly",xlab="Date")
mtext("C",side=3,line=-1,adj=-0.24,cex=1.5)
plot(y=anomaly[2,],x=time(he10_model),ylab="anomaly",xlab="Date")
mtext("D",side=3,line=-1,adj=-0.24,cex=1.5)
@

Trace plots to check on the convergence of \code{ibpf} are similar to those for other iterated filtering algorithms.
In Figure~\ref{fig:traces}, we show this for the Gaussian correlated random walk example.
In this case, we see replicable convergence among independent searches.

<<traces,fig.height=6.5, fig.width=6.5, out.width="6.5in", fig.cap = "Trace plots for the correlated Gaussian random walk example.">>=
ibpf_traces <- pomp::melt(lapply(ibpf_mle_searches,pomp:::traces_internal))
ibpf_traces$iteration <- as.numeric(ibpf_traces$iteration)
ggplot(ibpf_traces, aes(x=iteration,y=value,group=.L1,color=factor(.L1))) +
  geom_line() +
  guides(color="none") +
  facet_wrap(~variable,scales="free_y")
@
